---
title: "Spatial Analysis: 03"
author: "Tzai-Hung Wen"
date: '2025-03-10'
output:
  html_notebook:
    toc: true
    toc_depth: 6
    toc_float: true
---

# Spatial Intersection: Point-In-Polygon

```{r}
rm(list = ls())

library(sf)
library(tmap)

load("./data/Sample2.RData")
```

## 1.1 Tornado Damage Assessment 

### Mapping
```{r}
us_bg <- tm_shape(us_states_sf) + tm_polygons("grey90")
tornado <- tm_shape(torn_sf) + tm_dots(size = 0.04, shape = 1, fill_alpha = 0.5)
us_border <- tm_shape(us_states_sf) + tm_borders(col = "black") + tm_layout(frame = F) 
us_bg + tornado + us_border
```
### Checking CRS and data attribute
```{r}
st_crs(us_states_sf)
head(torn_sf)
summary(torn_sf)
st_geometry(torn_sf)
```

### Attribute selection
```{r}
index <- us_states_sf$STATE_NAME == "Texas" | us_states_sf$STATE_NAME == "New Mexico" | 
         us_states_sf$STATE_NAME == "Oklahoma" | us_states_sf$STATE_NAME == "Arkansas"  

AoI_sf <- us_states_sf[index,]
AoI_bg <- tm_shape(AoI_sf) + tm_polygons("grey90")
AoI_bg 
```

### Clip
```{r}
torn_clip_sf <- torn_sf[AoI_sf,]
torn_clip <- tm_shape(torn_clip_sf) + tm_dots(fill = "red", size = 0.04, fill_alpha = 0.5)
AoI_bg + torn_clip

head(torn_clip_sf)
```

### Point-In-Polygon: Using st_join()
```{r}
AoI_torn_sf <- st_join(torn_clip_sf, AoI_sf, join = st_within)
head(AoI_torn_sf)

AoI_torn_df <- as.data.frame(AoI_torn_sf)
head(AoI_torn_df)
summary(AoI_torn_df)
```

### Creating new data frame
```{r}
AoI_torn_df$STATE_NAME <- droplevels(AoI_torn_df$STATE_NAME) 
newdf<- data.frame(Name = AoI_torn_df$STATE_NAME,Damage= AoI_torn_sf$DAMAGE)
head(newdf)
newdf$Damage <- as.integer(newdf$Damage)
```

### Crosstab analysis
```{r}
count<- table(newdf$Damage, newdf$Name)
count

count2<- xtabs(~ newdf$Damage + newdf$Nam )
count2
```

## 1.2 Mapping Tornado Density 
```{r}
# (Counting Points in Polygons)

points_sf_joined <- st_join(torn_sf, us_states_sf, join = st_within)
head(points_sf_joined)

table1 <- table(points_sf_joined$STATE_NAME)

count_sf <- as.data.frame(table1)
colnames(count_sf) <- c("STATE_NAME","Counts")
head(count_sf)

library(dplyr)
us_states_sf<- left_join(us_states_sf, count_sf)
head(us_states_sf)

area1<-st_area(us_states_sf) # unit: foot

library(units)
area1<-set_units(area1, km^2)
us_states_sf$AREA1 <- area1
us_states_sf$Density <- us_states_sf$Counts / us_states_sf$AREA1
head(us_states_sf)

```
```{r}
library(pals) # collection of color palettes
plot(us_states_sf["Density"], breaks = "jenks", nbreaks = 6, pal=brewer.reds(6))

```

# Buffer zone
```{r}
st_crs(places_sf)
st_crs(places_sf)<- st_crs(blocks_sf)
head(places_sf)

block_bg <- tm_shape(blocks_sf) + tm_polygons("grey90")
places_pts <- tm_shape(places_sf) + tm_dots(fill = "red", size = 0.5)

block_bg + places_pts

places_buf_sf <- st_buffer(places_sf, dist = 3000)
summary(places_buf_sf)

places_buf <- tm_shape(places_buf_sf) + tm_polygons("yellow")
block_bg + places_buf + places_pts

places_bufU_sf <- st_union(places_buf_sf)
head(places_bufU_sf)

places_bufU <- tm_shape(places_bufU_sf) + tm_polygons("yellow")
block_bg + places_bufU + places_pts
```

# Distance Analysis

## checking attribute table
```{r}
head(blocks_sf)
```

## using st_centroid()
```{r}
blocks_center_sf <- st_centroid(blocks_sf)
blocks_center <- tm_shape(blocks_center_sf) + tm_dots(fill = "blue", size = 0.2)
block_bg +blocks_center
blocks_center + places_pts
```

## using st_distance()
```{r}
distance.matrix <- st_distance(blocks_center_sf, places_sf)
library(units)
distance.matrix<-set_units(distance.matrix, km)
```

## using apply()
```{r}
near_dist<- apply(distance.matrix, 1, mean)
near_dist[1]

xid<- which.min(near_dist)
near_dist[xid]
blocks_sf[xid,]
blocks_sf$POP1990[xid]

min_dist<- apply(distance.matrix, 1, min)
sel_blocks<- min_dist < 1  # < 1 km
sel_sf <- blocks_sf[sel_blocks,]
sel_map <- tm_shape(sel_sf) + tm_polygons("yellow")
block_bg + sel_map + places_pts
```

## using st_is_within_distance()
```{r}
d1<- set_units(1, km)
sel_blocks = st_is_within_distance(blocks_center_sf, places_sf, dist = d1)
class(sel_blocks)
sel_blocks<- lengths(sel_blocks) > 0
sel_sf <- blocks_sf[sel_blocks,]
sel_map <- tm_shape(sel_sf) + tm_polygons("yellow")
block_bg + sel_map + places_pts
```

## choose specific one place
```{r}
sel_blocks = st_is_within_distance(blocks_center_sf, places_sf[3,], dist = d1)
sel_blocks <- lengths(sel_blocks) > 0
sel_sf <- blocks_sf[sel_blocks,]
sel_map <- tm_shape(sel_sf) + tm_polygons("yellow")
block_bg + sel_map + places_pts

TOTAL_POP <- sum(sel_sf$POP1990)
TOTAL_WHITE <- sum(sel_sf$POP1990 * (sel_sf$P_WHITE/100) )
```

## accessibility analysis: using apply()
```{r}
eth<- as.data.frame(blocks_center_sf[,14:18])
eth<- as.matrix(eth[,1:5])
eth<- apply(eth, 2, function(x) (x*blocks_center_sf$POP1990))
colnames(eth)<- c("White","Black","Native","Asian","Other")
eth.access <- xtabs(eth~sel_blocks) 

eth.access <- as.data.frame(eth.access)
colnames(eth.access) <- c("Access","Eth5","Pop")

library(ggplot2)
ggplot(eth.access) + aes(fill=Access, y=Pop, x=Eth5) + 
  geom_bar(position="stack", stat="identity")
```

