---
title: "Spatial Analysis:02"
author: "Tzai-Hung Wen"
date: '2025-03-03'
output:
  html_notebook:
    toc: true
    toc_depth: 6
    toc_float: true
---

# Spatial Data Handling

```{r}
rm(list = ls()) # 清除任何現有的R物件之前執行批次工作的程式碼

library(sf)
library(tmap)
library(pals) # collection of color palettes
library(cartography) # classification methods

#setwd("C:/Wen_Files/SA_2024")
load("./data/Sample.RData")
```

# 0. sf format and coordinates
```{r}
class(blocks_sf)
head(blocks_sf)
```

## Extracing as a new layer

```{r}
new_sf <- blocks_sf[,6]
new2_sf <- blocks_sf[1:3,]
```

## Attribute table

```{r}
blocks_df <- as.data.frame(blocks_sf)
class(blocks_df)
head(blocks_df)
```

## Coordinate system

```{r}
st_crs(blocks_sf)
```
```{r}
st_crs(roads_sf)
```

```{r}
st_crs(roads_sf) <- st_crs(blocks_sf)
st_crs(roads_sf)
```
### Taiwan Coordinate System
```{r}
TWCOUNTY <- st_read("./data/TaiwanCounty.shp", options="ENCODING=utf-8")
st_crs(TWCOUNTY)
```

## Export/import shapefiles

```{r}
st_write(blocks_sf,"blocks.shp", delete_layer = TRUE)
```

#  1. Mapping Spatial Objects

## using plot

```{r}
plot(blocks_sf)
plot(blocks_sf["P_VACANT"], breaks = "jenks", nbreaks = 6)
#(blocks_sf$P_VACANT ?)

brewer.blues(6)
plot(blocks_sf["P_VACANT"], breaks = "jenks", nbreaks = 6, pal=brewer.blues(6))
```
## using tmap package

### quick tmap

```{r}
qtm(blocks_sf, fill="orange", style="natural")
# colors()
```
### Qick tmap: choropleth
```{r}
qtm(blocks_sf, fill="P_VACANT", fill.title="Vacant %", title="My Map 1")
```
### Qick tmap: bubble map
```{r}
qtm(blocks_sf, symbols.size="P_VACANT", symbols.title.size="Vacant %", title="My Bubble Map")
```
### Overlay multiple plots
```{r}
lyr1 <- qtm(blocks_sf, fill="P_VACANT", fill.title="Vacant %", title="My Map 1")
lyr_road <- tm_shape(roads_sf)+tm_lines(col="gray") # lines
lyr_crimes <- tm_shape(breach_sf)+tm_dots(col="red", size= 0.1) # points

# overlay multiple plots
lyr1 + lyr_road + lyr_crimes
```
### Showing multiple plots/maps

```{r}
library(grid)
# open a new plot page

grid.newpage()
# set up the layout
pushViewport(viewport(layout=grid.layout(1,2)))

# plot using the print command
print(lyr1, vp=viewport(layout.pos.col = 1))
print(lyr_road, vp=viewport(layout.pos.col = 2))

# dev.off() # close plotting device

```

# 2. Attribute Query

```{r}
index <- (blocks_sf$P_VACANT > 10)
newblocks_sf <- blocks_sf[index,]
lyr3 <- qtm(newblocks_sf, fill="red", title="Vacant > 10%", style="natural")
lyr_bg <- qtm(blocks_sf, fill="grey")
lyr_bg + lyr3
```

# 3. Calculating Fields

## create a field
```{r}
x <- st_area(blocks_sf) # unit: foot

library(units)
x2 <- set_units(x, km^2)
blocks_sf$AREA1 <- x2
colnames(blocks_sf)
```
## remove a field
```{r}
blocks_sf <- subset(blocks_sf, select = -c(AREA1))
colnames(blocks_sf)
```

## calculating values
```{r}
x2 <- set_units(x, km^2)
blocks_sf$AREA1 <- x2

blocks_sf$POPDEN <- blocks_sf$POP1990 / blocks_sf$AREA1

plot(blocks_sf["POPDEN"], breaks = "jenks", nbreaks = 6, pal= brewer.blues(6))
qtm(blocks_sf, fill="POPDEN", fill.title="Population Density", title="Popn Map", style="natural")
```
# 4. Detailed Settings for Mapping
```{r}
breakv<- getBreaks(v = blocks_sf$P_OWNEROCC, nclass = 6, method = "jenks")
# classification method:"fixed", "sd", "equal", "pretty", "quantile", "kmeans", 
# "hclust", "bclust", "fisher", "jenks", "dpih", "q6", "geom", "arith", "em", "msd" 

tm_shape(blocks_sf) +
  tm_polygons("P_OWNEROCC", title = "Owner Occ", palette = "-GnBu", 
              breaks = c(breakv), 
              legend.hist = T) +
  tm_scale_bar(width = 0.22) +
  tm_compass(position = c(0.8, 0.08)) +
  tm_layout(frame = F, title = "New Haven", 
            title.size = 2, title.position = c(0.55, "top"), 
            legend.hist.size = 0.5)
```


# 5. Statistical Plots

## Histogram
```{r}
hist(blocks_sf$P_VACANT, breaks = 40, col = "grey", 
     border = "red", 
     main = "The histogram of vacant property percentages", 
     xlab = "percentage vacant", xlim = c(0,40))
```
## using ggplot2: histogram
```{r}
library(ggplot2)
blocks_df <- as.data.frame(blocks_sf)
head(blocks_df)

plot1<- ggplot(blocks_df) + aes(P_VACANT) +
          geom_histogram(col = "red", fill = "grey", bins = 40) +
          xlab("percentage vacant") +
          labs(title = "The histogram of vacant property percentages")

plot1
```

## Boxplot
```{r}
boxplot(blocks_sf$P_WHIT, blocks_sf$P_BLACK, names=c("White", "Black"), 
        xlab="Race", ylab="Percentage")
```

## using ggplot2: boxplot
```{r}
library(reshape2)
block_prop <- melt(blocks_df[, c("P_WHITE", "P_BLACK", "P_AMERI_ES")])
head(block_prop)

plot2 <- ggplot(block_prop) + 
           aes(variable, value) +
           geom_boxplot()

plot2

```
## Showing map + plot

```{r}
grid.newpage()
pushViewport(viewport(layout=grid.layout(1,2)))
print(lyr1+lyr_crimes, vp=viewport(layout.pos.col = 1))
print(plot2, vp=viewport(layout.pos.col = 2))

#dev.off()

```

