---
title: "Spatial Analysis: 05"
author: "Tzai-Hung Wen"
date: '2025-03-31'
output:
  html_notebook:
    toc: true
    toc_depth: 6
    toc_float: true
---
# <h3> **Contents** </h3>

<h5>
*Using R package:aspace*

**Measures of Centrality** <br>
1.(Weighted) Mean center  <br>
2.Median center <br>
3.Central feature <br>

**Measures of Dispersion** <br>
1.Standard Distance <br>
2.Weighted Std. Distance <br>
3.Standard Deviational Ellipse <br> 

</h5>
 
# <h3> **0. Laoding Data** </h3>

```{r, message = FALSE}
rm(list = ls())
library(sf)
library(tmap)
library(aspace) #install.packages("aspace")
library(tidyverse)

schools_sf = st_read("./data/Schools.shp",options="ENCODING=BIG5")
head(schools_sf)
schools_lyr <- tm_shape(schools_sf)+tm_dots(col="red", size= 0.1) + tm_layout(frame = F)

```

# <h3> **1. Data Preparation** </h3>
<h5> (creating required fields)  </h5>
```{r}

# Generating no. of student in each school
schools_sf$Students<-as.integer(runif(424,1000,10000))

# Generating school type: cluster vs. isolation
for (i in 1: 424) {
  
  if (schools_sf$NEAR_DIST[i]< 500) {
    schools_sf$type[i]<- "Cluster"
  } else schools_sf$type[i]<- "Isolation"
  
}

index<- schools_sf$type == "Cluster"
school_cluster <- schools_sf[index,]

length(school_cluster)
schools.c_lyr <- tm_shape(school_cluster)+tm_dots(col="red", size= 0.1) + tm_layout(frame = F)

School_df <- data.frame(x=schools_sf$X_coor, y=schools_sf$Y_coor, 
                       type=schools_sf$type, students=schools_sf$Students)
```

# <h3> **2. Mean Center** </h3>
```{r}
# calc_mnc() Mean Centre Calculator
Mean.Center <- calc_mnc(id=1, weighted=FALSE, weights=NULL, points=School_df[,1:2])
W.Mean.Center <- calc_mnc(id=1, weighted=TRUE, weights=School_df$students, points=School_df[,1:2])

Mean.xcoor <- Mean.Center$LOCATIONS[2]
Mean.ycoor <- Mean.Center$LOCATIONS[3]

Mean.Center_sfg = st_point(c(Mean.xcoor, Mean.ycoor))
Mean.Center_sfc = st_sfc(Mean.Center_sfg)
Mean.Center_sf <- st_sf(Mean.Center_sfc)
head(Mean.Center_sf) 

# using pipe coding style
Mean.Center.Coor <- c(Mean.xcoor, Mean.ycoor)
Mean.Center_sf1 <- Mean.Center.Coor %>% st_point %>% st_sfc %>% st_sf 
class(Mean.Center_sf1)

st_crs(Mean.Center_sf1) <- st_crs(schools_sf)
Mean.Center_lyr <- tm_shape(Mean.Center_sf1)+tm_dots(fill="blue", size=1)

# Weighted Mean Center

W.Mean.xcoor <- W.Mean.Center$LOCATIONS[2]
W.Mean.ycoor <- W.Mean.Center$LOCATIONS[3]

W.Mean.Center.Coor <- c(W.Mean.xcoor, W.Mean.ycoor)
W.Mean.Center_sf <- W.Mean.Center.Coor %>% st_point %>% st_sfc %>% st_sf 
st_crs(W.Mean.Center_sf) <- st_crs(schools_sf)
W.Mean.Center_lyr <- tm_shape(W.Mean.Center_sf)+tm_dots(fill="green", size= 0.5)

schools_lyr + Mean.Center_lyr + W.Mean.Center_lyr
```


# <h3> **3. Median Center** </h3>
```{r}
# calc_mdc(): Median Centre Calculator
Median.Center <- calc_mdc(id=1,points=School_df[,1:2])

Median.xcoor <- Median.Center$LOCATIONS[2]
Median.ycoor <- Median.Center$LOCATIONS[3]

Median.Center.Coor <- c(Median.xcoor, Median.ycoor)
Median.Center_sf <- Median.Center.Coor %>% st_point %>% st_sfc %>% st_sf 
st_crs(Median.Center_sf) <- st_crs(schools_sf)
Median.Center_lyr <- tm_shape(Median.Center_sf)+tm_dots(fill = "green", size= 0.5)

### Compare Mean vs. Median
schools_lyr + Mean.Center_lyr + Median.Center_lyr
```


# <h3> **4. SDD: Standard Distance** </h3>
<h5>*using plot_sdd*</h5>
```{r}
# calc_sdd(): Calculate the Standard Distance Deviation (Standard Distance)
school.SDD <- calc_sdd(id=1, points=School_df[,1:2])
plot_sdd(school.SDD, centre.col="red")

```
<h5>*using tmap*</h5>
```{r}
center.x<- school.SDD$ATTRIBUTES$CENTRE.x
center.y<- school.SDD$ATTRIBUTES$CENTRE.y

center.coor <- c(center.x, center.y)
center_sf<- center.coor %>% st_point %>% st_sfc %>% st_sf 

st_crs(center_sf) <-  st_crs(schools_sf)

rad<- school.SDD$ATTRIBUTES$SDD.radius
SD_sf<- st_buffer(center_sf, rad)
SD_lyr <- tm_shape(SD_sf) + tm_borders(col = "blue")+tm_layout(frame = F)
Center_lyr <-  tm_shape(center_sf) + tm_dots(fill="blue", size= 0.5)
schools_lyr+ Center_lyr+ SD_lyr 

# weighted SDD
school.SDD2 <- calc_sdd(id=1, points = School_df[,1:2], 
                       weighted = TRUE, weights=School_df$students)
```



# <h3> **5. SDE:Standard Deviational Ellipse** </h3>

<h5>*using plot_sde*</h5>
```{r}
## SDE Center: sdeatt
# calc_sde() Calculate the Standard Deviation Ellipse
school.SDE<- calc_sde(id=1, points=School_df[,1:2])
plot_sde(school.SDE)
```

<h5>*using tmap*</h5>

```{r}
## Creating SDE center

SDE.center.x<- school.SDE$ATTRIBUTES$CENTRE.x
SDE.center.y<- school.SDE$ATTRIBUTES$CENTRE.y
SDE.center.coor <- c(SDE.center.x, SDE.center.y)
SDE.center_sf <- SDE.center.coor %>% st_point %>% st_sfc %>% st_sf 
st_crs(SDE.center_sf) <-  st_crs(schools_sf)
SDE.center_lyr <- tm_shape(SDE.center_sf) + tm_dots(fill = "blue", size=0.5)+tm_layout(frame = F)

## Creating SDE Polygon

xcoor <- school.SDE$LOCATIONS$x
ycoor <- school.SDE$LOCATIONS$y
xy = data.frame(x=xcoor, y=ycoor)
xys = st_as_sf(xy, coords=c("x","y"))
st_crs(xys)<-st_crs(schools_sf)

SDE_sfc <- st_cast(st_combine(xys),"POLYGON")
SDE_sf <- st_sf(SDE_sfc)
SDE_lyr <- tm_shape(SDE_sf) + tm_borders(col = "blue") +tm_layout(frame = F)

schools_lyr + SDE.center_lyr + SDE_lyr
```

# <h3> **6. Central Feature** </h3>
```{r}
school.CF <- calc_cf(id=1, points=School_df[,1:2])

CF.x<- school.CF$ATTRIBUTES$CF.x
CF.y<- school.CF$ATTRIBUTES$CF.y
CF.coor <- c(CF.x, CF.y)
CF_sf <- CF.coor %>% st_point %>% st_sfc %>% st_sf 
st_crs(CF_sf) <-  st_crs(schools_sf)
CF_lyr <- tm_shape(CF_sf) + tm_dots(fill = "blue", size= 0.5) +tm_layout(frame = F)
schools_lyr + CF_lyr
```


# <h3> **7. Mean Centers by Grouping Attribubtes** </h3>
```{r}
type <- schools_sf$type
newid <- unique(type)

schools_lyr2 <- tm_shape(schools_sf) + 
                tm_dots(col="type", palette=c(Cluster='red', Isolation='orange'), size= 0.3) +
                tm_layout(frame = F)

schools_lyr2


xx<-vector(); yy<-vector(); ctype<-vector()

for (i in 1:2){
  index<-(type == newid[i])
  newschool<-schools_sf[index,]
  xcoor<-newschool$X_coor
  ycoor<-newschool$Y_coor
  newschool.mc <- calc_mnc(id=1, points=cbind(xcoor, ycoor))
  
  xx[i]<-newschool.mc$LOCATIONS[2]
  yy[i]<-newschool.mc$LOCATIONS[3]
  ctype[i]<-newid[i]
}

newcenterxy <- data.frame(xx,yy, ctype)
New_sf <- st_as_sf(newcenterxy , coords=c("xx","yy"))
st_crs(New_sf) <-  st_crs(schools_sf)

New_lyr <- tm_shape(New_sf) + 
           tm_dots(col="ctype", palette=c(Cluster='blue', Isolation='cyan'), size= 0.5) +
           tm_layout(frame = F) 


schools_lyr2 + New_lyr 
```

